Modelling state-switching behaviour in functional data with hidden Markov models

### 2020-09-04 Report/Sections/simulation.tex
Update report- Evan Edits

To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT. The parameters used to generate the data are loosely based on those estimated in the case study (see able \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states is set to $N^*=2$ and $\Ztwo_{t,t^*}$ is set to a scalar. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.

\subsection{Data simulation}

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
The observations $Z^*$ are generated using the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
Recall that $\Zone_{t,t^*}$ follows a CarHMM and is Normally distributed while $\Ztwo_{t,t^*}$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.

The two dive types differ in that dives of type 1 are much shorter on average than dives of type 2. The two subdive types differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher in subdive state 2 than subdive state 1. This corresponds to much more vigorous and variable periodic behaviour in the acceleration data.

@@

As expected, since it is the generating model, the full CarHHMM-DFT is the best-performing model. Using the forward-backward algorithm, the CarHHMM-DFT decodes the correct subdive type given the fine-scale data with an average confidence of $1 - 10^{-12}$ across all simulations. In other words, it predicts the \textit{incorrect} subdive state with an average probability of $10^{-12}$. For dive type, the average confidence was above 0.9 (see Table \ref{table:accuracy}). All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased (see section 2 of the supplementary material). Even still, these biases are the least severe for the CarHHMM-DFT among the four models. The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error. See Tables 6-9 of the supplementary material for more information.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. However, the HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_A^*$. For both subdive types, $\hat \sigma_A^*$ greatly overestimates the true standard deviation. The estimated standard errors of $\hat \mu_A^*$ and $\hat \sigma_A^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Table 7 of the supplementary material). This implies that the variance of acceleration can be overestimated and the standard error of the parameter estimates can be underestimated when auto-correlation is ignored. These findings are consistent with the results from the case study, where the HHMM-DFT produced higher variance estimates and narrower standard error estimates than the CarHHMM-DFT.

The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types. This result is consistent with expectations because the CarHHMM does not have access to the ``wiggles" of the fine-scale process, which are the most informative observations when determining subdive type. The CarHHMM is also much more likely to misclassify subdive type 1 than subdive type 2 (see Table \ref{table:accuracy}). The CarHHMM is also the worst of the four at estimating parameters in part due to its relatively poor decoding accuracy. The fine-scale probability transition matrices associated with dive type 2 ($\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$) are especially poor estimates (see Table 9 and Figure 33 of the supplementary material).

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, it lacks any hierarchical structure, so it is unable to estimate dive types. The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of \textit{all} dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds. See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.

>>>>

To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data has equal numbers of short and long dives in expectation, the number subdive states is set to $N^*=2$, and $\Ztwo_{t,t^*}$ is set to a scalar. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.

\subsection{Data simulation}
\label{subsec:data_simulation}

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which each represent a simulated killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\theta$ are
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
For each dive and subdive segment, we simulate the fine-scale transformed observations $Z^*_t \equiv \left\{Z^*_{t,1}, \ldots, \right\}$, where $Z^*_{t,t^*} = \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$, using the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
Recall that $\Zone_{t,t^*}$ follows a CarHMM and is Normally distributed while $\Ztwo_{t,t^*}$ follows an HMM and is Gamma distributed. It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.

The two simulated dive types differ in that dives of type 1 are much shorter on average (20 seconds) than dives of type 2 (80 seconds). The two subdive types differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher in subdive state 2 than subdive state 1. This corresponds to much more vigorous and variable periodic behaviour in the acceleration data.

@@

The full CarHHMM-DFT is the best-performing model of the four candidates, which is to be expected as the generating model. If $X^*_{t,t^*}$ is the unknown subdive state and $x^*_{t,t^*}$ is the known true subdive state for dive $t$ and window $t^*$, then using the forward-backward algorithm we find that $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ is $(1 - 10^{-12})$ on average across all simulations, dives, and windows. We refer to this quantity as the average accuracy of the model. For dive type, this average accuracy was above 0.9 (see Table \ref{table:accuracy}). All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases. In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are no more severe for the CarHHMM-DFT than the other three models. The empirical standard error of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard error. See Tables 6-9 of the supplementary material for more information.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its average accuracy and parameter bias. However, the HHMM-DFT differs from the CarHHMM-DFT in its estimates of $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$, both of which are greatly overestimated. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material). These results imply that estimated standard errors of parameter estimates are too small when using the HHMM-DFT since auto-correlation is ignored. This finding is consistent with the results from the case study, where the HHMM-DFT produced higher estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.

The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average accuracy below $0.9$ for both dive and subdive types. This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct observation when determining subdive type. The CarHHMM is also much more likely to misclassify subdive type 1 than subdive type 2 (see Table \ref{table:accuracy}). In addition to its poor average accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters, with the fine-scale probability transition matrices associated with dive type 2 ($\hat \theta^{*(2)}$ and $\hat \Gamma^{*(2)}$) being especially poor estimates (see Table 9 and Figure 33 of the supplementary material).

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model is unable to estimate dive types as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by approximately five seconds. See Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive types and subdive states for each model.

### 2020-09-04 Report/main.tex
Commit Before Andrew Access
Updates from Overleaf
Clean up report and delete old files
Update report- Evan Edits

% Initial and revised submissions should be 12 point; this will be removed in the final version.
\documentclass[12pt]{TD-CJS}

@@

% User-defined commands go here
\renewcommand{\eqref}[1]{(\ref{#1})}
\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mt}[1]{\mathrm{#1}}
\newcommand{\rv}{random variable}
\newcommand{\newblock}{}
\bibliographystyle{abbrvnat}

@@

Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields. These data often exhibit complicated temporal dependence structure which cannot be modelled using common methods of Functional Data Analysis (FDA). Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs), but many intricate fine-scale processes violate basic assumptions of HMMs.
We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at fine scales. 
%We also provide several examples of viable fine-scale models.
As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. 
Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing models.

@@

%%%%%%%%%%%%
%\vfill\eject
%{\centerline{\Large{Notes from Nancy}}}
%\input{Sections/0627NH}
%\vfill\eject
%%%%%%%%%%%%%
%
\section{Introduction}
\input Sections/intro.tex
%
\section{Models and parameter estimation}
\label{sec:models}
\input Sections/models.tex
%
\section{Killer Whale Case Study}
\label{sec:data}
\input Sections/results.tex
%
\section{Simulation Study}
\label{sec:sim_study}
\input Sections/simulation.tex
%
%\section{Case Study Results}
%\label{sec:case_study}
%\input Sections/results.tex
%%
\section{Discussion}
\input Sections/discussion.tex
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{Figures and Tables}
\input Sections/figures.tex
%
%\CJShistory

>>>>

% Initial and revised submissions should be 12 point; this will be removed in the final version.
\documentclass[12pt]{TD-CJS}
%\documentclass{TD-CJS}

@@

% User-defined commands go here
\renewcommand{\eqref}[1]{(\ref{#1})}
\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mt}[1]{\mathrm{#1}}
\newcommand{\rv}{random variable}
\newcommand{\newblock}{}
\bibliographystyle{apalike}

@@

Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields. These data often exhibit complicated temporal dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA). Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs.
We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at fine scales. 
%We also provide several examples of viable fine-scale models.
As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. 
Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.

@@

%%%%%%%%%%%%
%\vfill\eject
%{\centerline{\Large{Notes from Nancy}}}
%\input{Sections/0627NH}
%\vfill\eject
%%%%%%%%%%%%%
%
\section{INTRODUCTION}
\input Sections/intro.tex
%
\section{MODELS AND PARAMETER ESTIMATION}
\label{sec:models}
\input Sections/models.tex
%
\section{KILLER WHALE CASE STUDY}
\label{sec:data}
\input Sections/results.tex
%
\section{SIMULATION STUDY}
\label{sec:sim_study}
\input Sections/simulation.tex
%
%\section{Case Study Results}
%\label{sec:case_study}
%\input Sections/results.tex
%%
\section{DISCUSSION}
\input Sections/discussion.tex
%
%\newpage
%\input Sections/bibliography.tex
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{FIGURES AND TABLES}
\input Sections/figures.tex
%
%\CJShistory
